# Chapter 4: Clustering and classification 

*In this chapter, we will practice clustering and classification with example data.*


We will start by reading in the Boston data from the MASS package. The data consists of 506 rows and 14 columns. It has both numeric and integer variables. It reflects Housing Values in Suburbs of Boston. 


```{r}
# load library
library(MASS)

# load the data
data("Boston")

# explore the dataset
str(Boston)
dim(Boston)
```

Most of the variables are skewed to some direction. Only a few, like rm and medv are realatively normally distributed. 
The strongest correlations are between the variables rad and tax (neg), dis and age (pos), dis and nox (pos), dis and indus, lstat and medv (pos).


```{r}
# load library
library(corrplot)
library(dplyr)
require(reshape2)

# convert wide to long
melt.boston <- melt(Boston)

# plot distributions 
ggplot(data = melt.boston, aes(x = value)) + 
stat_density() + 
facet_wrap(~variable, scales = "free")

# plot simple correlation plot 
cor_matrix <- cor(Boston) %>% round(2)
corrplot(cor_matrix, method = "circle", type = "upper", cl.pos = "b", tl.pos = "d", tl.cex = 0.6)
 
# show summaries of the variables 
summary(Boston)

```

Lets standardize all the values (mean = 0), and create a categorical variable "crime". Let's also divide the dataset to train and test sets, so that 80% of the data belongs to the train set. 

```{r}
# center and standardize variables
boston_scaled <- scale(Boston)

# print summaries of the scaled variables
summary(boston_scaled) # note that now all means are 0 as supposed to 

# change the object to data frame
boston_scaled_df <- as.data.frame(boston_scaled)

# create a quantile vector of crim
bins <- quantile(boston_scaled_df$crim)

# create a categorical variable 'crime'
crime <- cut(boston_scaled_df$crim, breaks = bins, include.lowest = TRUE, label = c("low", "med_low", "med_high", "high"))

# remove original crim from the dataset
boston_scaled_df <- dplyr::select(boston_scaled_df, -crim)

# add the new categorical value to scaled data
boston_scaled_df <- data.frame(boston_scaled_df, crime)

# number of rows in the Boston dataset 
n <- nrow(boston_scaled_df)

# choose randomly 80% of the rows
ind <- sample(n,  size = n * 0.8)

# create train set
train <- boston_scaled_df[ind,]

# create test set 
test <- boston_scaled_df[-ind,]

```

Next, we will fit a linear discriminant analysis on the train set, and use the categorical crime rate as the target variable and all the other variables as predictor variables. 

```{r}
# fit a linear discriminant analysis
lda.fit <- lda(crime ~ . , data = train)

# the function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "red", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  graphics::arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

# target classes as numeric
classes <- as.numeric(train$crime)

# plot the lda results (select both lines and execute them at the same time!)
plot(lda.fit, dimen = 2)
lda.arrows(lda.fit, myscale = 1)
```
Let's save the crime categories from the test set and remove the categorical crime variable from the test dataset.
Next, we will predict the classes with the LDA model on the test data. The cross tabulated results show that a relatively small amount of the observations were falsely categorized by the model. For example, from med_low's, only 8 observations were altogether falsely categorized, whereas 17 was correclty categorized. 

```{r}
# save the classes from test data
classes <- test$crime

# remove the crime variable from test data
test <- dplyr::select(test, -crime)

# predict classes with test data
lda.pred <- predict(lda.fit, newdata = test)

# cross tabulate the results
table(correct = classes, predicted = lda.pred$class)
```

Lets's reload the Boston dataset, standardize it and calculate the distances between the observations. 

After this, we will run k-means algorithm on the dataset and investigate what is the optimal number of clusters. According to the visualization, the optimal number is 2 (i.e. the point when the total WCSS drops radically). 

```{r}
# load the data
data("Boston")

# scale the Boston dataset 
boston_scaled <- scale(Boston)

# calculate euclidean distances
dist_eu <- dist(boston_scaled)

# look at the summary of the distances
summary(dist_eu)

# set seed to get reproducable results 
set.seed(123)

# determine the number of clusters
k_max <- 10

# calculate the total within sum of squares
twcss <- sapply(1:k_max, function(k){kmeans(boston_scaled, k)$tot.withinss})

# visualize the results
qplot(x = 1:k_max, y = twcss, geom = 'line') # the optimal number of clusters seems to be around 2

# k-means clustering
km <- kmeans(boston_scaled, centers = 2)

# plot the Boston dataset with clusters
pairs(boston_scaled, col = km$cluster)

```

*Bonus: Perform k-means on the original Boston data with some reasonable number of clusters (> 2). Remember to standardize the dataset. Then perform LDA using the clusters as target classes. Include all the variables in the Boston data in the LDA model. Visualize the results with a biplot (include arrows representing the relationships of the original variables to the LDA solution). Interpret the results. Which variables are the most influential linear separators for the clusters? (0-2 points to compensate any loss of points from the above exercises)*

*Super-Bonus: Run the code below for the (scaled) train data that you used to fit the LDA. The code creates a matrix product, which is a projection of the data points.*