# Chapter 6: Analysis of longitudinal data 

*In this chapter, we will practice analysing longitudinal data.*

*1. Implement the analyses of Chapter 8 of MABS, using the R codes of Exercise Set 6: Meet and Repeat: PART I but using the RATS data (from Chapter 9 and Meet and Repeat: PART II). (0-7 points: 0-4 points for graphs or analysis results + 0-3 points for their interpretations)*

```{r}
# read in data 
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = T)

# convert categoricals to factors in RATS
RATS$ID <- factor(RATS$ID)
RATS$Group <- factor(RATS$Group)

# convert RATS from wide to long 
RATSL <- pivot_longer(RATS, cols = -c(ID, Group), 
                      names_to = "groups",
                      values_to = "wd") %>% 
  mutate(Time = as.integer(substr(groups, 3, 4))) %>%
  arrange(Time) # order by time variable
```

*2. Implement the analyses of Chapter 9 of MABS, using the R codes of Exercise Set 6: Meet and Repeat: PART II, but using the BPRS data (from Chapter 8 and Meet and Repeat: PART I). (0-8 points: 0-4 points for graphs or analysis results + 0-4 points for their interpretations)*

```{r}
# read in data 
BPRS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", sep =" ", header = T)

# convert categoricals to factors in BPRS
BPRS$treatment <- factor(BPRS$treatment)
BPRS$subject <- factor(BPRS$subject)

# convert BPRS from wide to long 
BPRSL <-  pivot_longer(BPRS, cols = -c(treatment, subject),
                       names_to = "weeks", values_to = "bprs") %>%
  mutate(week = as.integer(substr(weeks, 5, 6))) %>% 
  arrange(weeks) # order by weeks variable
```

BPRS data, in which 40 male subjects were randomly assigned to one of two treatment groups and each subject was rated on the brief psychiatric rating scale (BPRS) measured before treatment began (week 0) and then at weekly intervals for eight weeks. The BPRS assesses the level of 18 symptom constructs such as hostility, suspiciousness, hallucinations and grandiosity; each of these is rated from one (not present) to seven (extremely severe). The scale is used to evaluate patients suspected of having schizophrenia.

*Feel free with the graphs and analyses!*